A generalization of the Bradley-Terry model for draws in chess

### 2020-10-04 inst/jebo_chess_revision2.tex
change some text from blue to darkgreen to match the rebuttal

\begin{abstract}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  \textcolor{blue}{I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-6}$).}
\end{abstract}

@@

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}
\newcommand{\Prob}{\operatorname{Prob}}

@@

\definecolor{DarkGreen}{RGB}{14,159,4}

@@

The present work details an approach to chess analysis based on
reification of entities representing proclivity to draw, first-mover
(or white player) advantage, and \textcolor{blue}{the additional
  playing strength conferred by previous collusive draws.  Likelihood
  functions for the players' and reified entities' strengths are
  defined using only specific tournament outcomes.}  Many hypotheses
of interest to the chess community may be analysed straightforwardly.

>>>>

\begin{abstract}
\definecolor{DarkGreen}{RGB}{14,159,4}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  \textcolor{DarkGreen}{I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-6}$).}
\end{abstract}

@@

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}
\newcommand{\Prob}{\operatorname{Prob}}
\definecolor{DarkGreen}{RGB}{14,159,4}

@@

@@

The present work details an approach to chess analysis based on
reification of entities representing proclivity to draw, first-mover
(or white player) advantage, and \textcolor{DarkGreen}{the additional
  playing strength conferred by previous collusive draws.  Likelihood
  functions for the players' and reified entities' strengths are
  defined using only specific tournament outcomes.}  Many hypotheses
of interest to the chess community may be analysed straightforwardly.
